On exponential sums arising from the classical doubling method

### 2025-02-09 paper.tex
Rephrasing
Writing the Hasse--Davenport relation for our case

\subsubsection{Split torus computation II}
We use the results from the previous section to compute the doubling method Gauss sum for a torus of the form $\multiplicativegroup{\finiteFieldExtension{k}}$. Let $\alpha \colon \multiplicativegroup{\finiteFieldExtension{k}} \to \multiplicativegroup{\cComplex}$ and $\chi \colon \multiplicativegroup{\finiteField} \to \multiplicativegroup{\cComplex}$ be characters such that $\chi^2 \ne 1$. Our goal is to compute $$\sum_{\substack{x \in \multiplicativegroup{\finiteFieldExtension{k}}\\
		x \ne 1}} \alpha \left(x\right) \chi\left(\FieldNorm{k}{1}\left(x - 1\right)\right) \chi\left(\FieldNorm{k}{1}\left(x^{-1} - 1\right)\right).$$
This can be rewritten as
$$\chi\left(-1\right)^k \sum_{\substack{x \in \multiplicativegroup{\finiteFieldExtension{k}}\\
		x \ne 1}} \alpha \left(x\right) \chi^{-1}\left(\FieldNorm{k}{1}\left(x\right)\right) \chi^2\left(\FieldNorm{k}{1}\left(x - 1\right)\right).$$
By the previous section, this equals
$$- q^{\frac{k}{2}} \alpha\left(-1\right)  \GaussSumCharacter{\alpha^{-1}}{\chi^{-1}}{\fieldCharacter_k} \GaussSumCharacter{\alpha}{\chi^{-1}}{\fieldCharacter_k} \tau\left(\chi^2, \fieldCharacter\right)^k.$$

@@

As usual, we rewrite this sum as follows
$$q^{-2m} \sum_{z \in \finiteFieldExtension{2m}} \sum_{y \in \multiplicativegroup{\finiteFieldExtension{2m}}} \sum_{x \in \finiteFieldExtension{2m}^1} \theta \left(x\right) \chi\left(\FieldNorm{2m}{1}\left(y\right)\right) \fieldCharacter_{2m}\left(z\left(y-x+1\right)\right).$$
If $z=0$ then the sum over $y$ will vanish. Hence we can reduce the sum to $z \in \multiplicativegroup{\finiteFieldExtension{2m}}$. Replacing $y$ with $z^{-1} y$ we get the sum
$$-q^{-m} \tau\left(\chi \circ \FieldNorm{2m}{1}, \fieldCharacter_{2m}\right) \sum_{z \in \multiplicativegroup{\finiteFieldExtension{2m}}} \chi^{-1}\left(\FieldNorm{2m}{1}\left(z\right)\right) \fieldCharacter_{2m}\left(z\right)  \sum_{x \in \finiteFieldExtension{2m}^1} \theta \left(-x\right) \fieldCharacter_{2m}\left(xz\right).$$
To proceed, we use the Hilbert 90 map $\multiplicativegroup{\finiteFieldExtension{2m}} \to \finiteFieldExtension{2m}^1$ given by $t \mapsto t^{1 - q^m}$. This map is surjective, and its kernel is $\multiplicativegroup{\finiteFieldExtension{m}}$. Define $\theta' \colon \multiplicativegroup{\finiteFieldExtension{2m}} \to \multiplicativegroup{\cComplex}$ by $\theta'\left(t\right) = \theta\left(t^{1-q^m}\right)$. Replacing the sum over $x \in \finiteFieldExtension{2m}^1$ with a sum over $t \in \multiplicativegroup{\finiteFieldExtension{2m}}$, we get \begin{align*}
	& \sum_{z \in \multiplicativegroup{\finiteFieldExtension{2m}}} \chi^{-1}\left(\FieldNorm{2m}{1}\left(z\right)\right) \fieldCharacter_{2m}\left(z\right) \sum_{x \in \finiteFieldExtension{2m}^1} \theta \left(x\right) \fieldCharacter_{2m}\left(xz\right) \\
	= & \frac{1}{q^m-1}\sum_{z \in \multiplicativegroup{\finiteFieldExtension{2m}}} \sum_{t \in \multiplicativegroup{\finiteFieldExtension{2m}}} \chi^{-1}\left(\FieldNorm{2m}{1}\left(z\right)\right) \fieldCharacter_{2m}\left(z^{q^m}\right) \theta' \left(t\right) \fieldCharacter_{2m}\left(t^{1-q^m} z\right).
\end{align*}
Replacing $z$ with $t^{q^m} z$, this becomes
\begin{align*}
	\frac{1}{q^m-1}\sum_{t,z \in \multiplicativegroup{\finiteFieldExtension{2m}}} \chi^{-1}\left(\FieldNorm{2m}{1}\left(z\right)\right) \chi^{-1}\left(\FieldNorm{2m}{1}\left(t\right)\right) \theta' \left(t\right) \fieldCharacter_{2m}\left(\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right) t\right).
\end{align*}
Since $\chi \circ \FieldNorm{2m}{1} \ne \theta'$, if $\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right) = 0$, we have that the inner sum over $t$ is zero. Hence, we may replace the sum over $z$ such that $\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right) \ne 0$. Using the fact that $\theta'$ is trivial on $\multiplicativegroup{\finiteFieldExtension{m}}$, we have after replacing variables $t \mapsto \frac{t}{\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right)}$,
\begin{align*}
	\frac{1}{q^m-1}\sum_{\substack{t,z \in \multiplicativegroup{\finiteFieldExtension{2m}}\\
	\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right) \ne 0}} \chi^{-1}\left(\FieldNorm{2m}{1}\left(\frac{z}{\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right) }\right)\right) \chi^{-1}\left(\FieldNorm{2m}{1}\left(t\right)\right) \theta' \left(t\right) \fieldCharacter_{2m}\left(t\right).
\end{align*}
By the appendix, this equals
\begin{align*}
	q^{\frac{m}{2}} \tau\left(\chi^{-1} \circ \FieldNorm{m}{1}, \fieldCharacter_m\right)^2 \tau\left(\chi^{2} \circ \FieldNorm{m}{1}, \fieldCharacter_m\right) \left(\sum_{t \in \multiplicativegroup{\finiteFieldExtension{2m}}} \chi^{-1}\left(\FieldNorm{2m}{1}\left(t\right)\right) \theta' \left(t\right) \fieldCharacter_{2m}\left(t\right)\right).
\end{align*}

>>>>

\subsubsection{Split torus computation II}
We use the results from the previous section to compute the doubling method Gauss sum for a torus of the form $\multiplicativegroup{\finiteFieldExtension{k}}$. Let $\alpha \colon \multiplicativegroup{\finiteFieldExtension{k}} \to \multiplicativegroup{\cComplex}$ and $\chi \colon \multiplicativegroup{\finiteField} \to \multiplicativegroup{\cComplex}$ be characters such that $\chi^2 \ne 1$. Our goal is to compute $$\sum_{\substack{x \in \multiplicativegroup{\finiteFieldExtension{k}}\\
		x \ne 1}} \alpha \left(x\right) \chi\left(\FieldNorm{k}{1}\left(x - 1\right)\right) \chi\left(\FieldNorm{k}{1}\left(x^{-1} - 1\right)\right).$$
This can be rewritten as
$$\chi\left(-1\right)^k \sum_{\substack{x \in \multiplicativegroup{\finiteFieldExtension{k}}\\
		x \ne 1}} \alpha \left(x\right) \chi^{-1}\left(\FieldNorm{k}{1}\left(x\right)\right) \chi^2\left(\FieldNorm{k}{1}\left(x - 1\right)\right).$$
By the previous section, this equals
\begin{equation}\label{eq:identity-for-multiplicative-group-of-f-k-gauss-sum}
	-q^{\frac{k}{2}} \alpha\left(-1\right)  \GaussSumCharacter{\alpha^{-1}}{\chi^{-1}}{\fieldCharacter_k} \GaussSumCharacter{\alpha}{\chi^{-1}}{\fieldCharacter_k} \tau\left(\chi^2, \fieldCharacter\right)^k.
\end{equation}

Notice that if $k = md$ and $\alpha = \beta \circ \FieldNorm{k}{m}$ where $\beta \colon \multiplicativegroup{\finiteFieldExtension{m}} \to \multiplicativegroup{\cComplex}$ is a character, then \eqref{eq:identity-for-multiplicative-group-of-f-k-gauss-sum} equals
$$-\left(q^{\frac{m}{2}} \GaussSumCharacter{\beta^{-1}}{\chi^{-1}}{\fieldCharacter_m} \GaussSumCharacter{\beta}{\chi^{-1}}{\fieldCharacter_m} \tau\left(\chi^2, \fieldCharacter\right)^m\right)^{d}.$$

@@

As usual, we rewrite this sum as follows
$$q^{-2m} \sum_{z \in \finiteFieldExtension{2m}} \sum_{y \in \multiplicativegroup{\finiteFieldExtension{2m}}} \sum_{x \in \finiteFieldExtension{2m}^1} \theta \left(x\right) \chi\left(\FieldNorm{2m}{1}\left(y\right)\right) \fieldCharacter_{2m}\left(z\left(y-x+1\right)\right).$$
If $z=0$ then the sum over $y$ will vanish. Hence we can reduce the sum to $z \in \multiplicativegroup{\finiteFieldExtension{2m}}$. Replacing $y$ with $z^{-1} y$ we get the sum
$$-q^{-m} \tau\left(\chi \circ \FieldNorm{2m}{1}, \fieldCharacter_{2m}\right) \sum_{z \in \multiplicativegroup{\finiteFieldExtension{2m}}} \chi^{-1}\left(\FieldNorm{2m}{1}\left(z\right)\right) \fieldCharacter_{2m}\left(z\right)  \sum_{x \in \finiteFieldExtension{2m}^1} \theta \left(-x\right) \fieldCharacter_{2m}\left(xz\right).$$
To proceed, we use the Hilbert 90 map $\multiplicativegroup{\finiteFieldExtension{2m}} \to \finiteFieldExtension{2m}^1$ given by $t \mapsto t^{1 - q^m}$. This map is surjective, and its kernel is $\multiplicativegroup{\finiteFieldExtension{m}}$. Define $\theta' \colon \multiplicativegroup{\finiteFieldExtension{2m}} \to \multiplicativegroup{\cComplex}$ by $\theta'\left(t\right) = \theta\left(t^{1-q^m}\right)$. Replacing the sum over $x \in \finiteFieldExtension{2m}^1$ with a sum over $t \in \multiplicativegroup{\finiteFieldExtension{2m}}$, we get \begin{align*}
	& \sum_{z \in \multiplicativegroup{\finiteFieldExtension{2m}}} \chi^{-1}\left(\FieldNorm{2m}{1}\left(z\right)\right) \fieldCharacter_{2m}\left(z\right) \sum_{x \in \finiteFieldExtension{2m}^1} \theta \left(x\right) \fieldCharacter_{2m}\left(xz\right) \\
	= & \frac{1}{q^m-1}\sum_{z \in \multiplicativegroup{\finiteFieldExtension{2m}}} \sum_{t \in \multiplicativegroup{\finiteFieldExtension{2m}}} \chi^{-1}\left(\FieldNorm{2m}{1}\left(z\right)\right) \fieldCharacter_{2m}\left(z^{q^m}\right) \theta' \left(t\right) \fieldCharacter_{2m}\left(t^{1-q^m} z\right).
\end{align*}
Replacing $z$ with $t^{q^m} z$, this becomes
\begin{align*}
	\frac{1}{q^m-1}\sum_{t,z \in \multiplicativegroup{\finiteFieldExtension{2m}}} \chi^{-1}\left(\FieldNorm{2m}{1}\left(z\right)\right) \chi^{-1}\left(\FieldNorm{2m}{1}\left(t\right)\right) \theta' \left(t\right) \fieldCharacter_{2m}\left(\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right) t\right).
\end{align*}
Since $\chi \circ \FieldNorm{2m}{1} \ne \theta'$, if $\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right) = 0$, we have that the inner sum over $t$ is zero. Hence, we may reduce the sum over $z$ to $z$ such that $\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right) \ne 0$. Using the fact that $\theta'$ is trivial on $\multiplicativegroup{\finiteFieldExtension{m}}$, we have after replacing variables $t \mapsto \frac{t}{\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right)}$,
\begin{align*}
	\frac{1}{q^m-1}\sum_{\substack{t,z \in \multiplicativegroup{\finiteFieldExtension{2m}}\\
	\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right) \ne 0}} \chi^{-1}\left(\FieldNorm{2m}{1}\left(\frac{z}{\trace_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{m}}\left(z\right) }\right)\right) \chi^{-1}\left(\FieldNorm{2m}{1}\left(t\right)\right) \theta' \left(t\right) \fieldCharacter_{2m}\left(t\right).
\end{align*}
By the appendix, this equals
\begin{align*}
	q^{\frac{m}{2}} \tau\left(\chi^{-1} \circ \FieldNorm{m}{1}, \fieldCharacter_m\right)^2 \tau\left(\chi^{2} \circ \FieldNorm{m}{1}, \fieldCharacter_m\right) \left(\sum_{t \in \multiplicativegroup{\finiteFieldExtension{2m}}} \chi^{-1}\left(\FieldNorm{2m}{1}\left(t\right)\right) \theta' \left(t\right) \fieldCharacter_{2m}\left(t\right)\right).
\end{align*}
